---
title: "Learn R from MarinStatsLectures"
output: html_notebook
---

This is an [R Youtube Lectures](https://www.youtube.com/watch?v=riONFzJdXcs&list=PLqzoL9-eJTNBDdKgJgJzaQcY6OXmsXAHU) Notebook. 

### 1.1 RStudio
```{r}
x<-1:5
y<-6:10
plot(x,y)
```

### 1.2 Install R and RStudio

Download [R](http://www.r-project.org) first.

Then install [RStudio] 

### 1.3 Getting Started


0:00:20 How to assign values to an object in R: 2 different ways
```{r}
y=12
x<- 11
print(y)
x
```
easy to overwrite.

0:01:46 How to use "ls" command to see what is stored in R (memory)
```{r}
ls()
```

0:01:57 How  remove an object using "rm" command
```{r}
rm(y)
y
```

0:02:52 How to assign character values to objects in R
```{r}
myName<-"Eric Q"
myName
```

### 1.4 Vectors, Matrices


0:00:29 how to create vectors in R for both numbers or objects using the "c" or "concatenate" command
```{r}
x1 <- c(1,3,4,7,9)
x1
gender <- c("male","female")
gender

```

0:01:09 how to create a sequence of integer values in R using the colon (:)
```{r}
x2 <- c(2:7)
x2
x2[2]
x2[-2]
x2[c(1,3)]
x2[c(1:3)]
```

0:01:20 how to use "seq" command in R to create sequences
```{r}
seq(from=1,to=7,by=2)
seq(1,10,3)

```

0:01:55 how to use "rep" command in R to create a vector of repeated numbers or characters
```{r}
rep(1, times=10)
rep("Eric",5)
rep(1:3,5)
```

0:03:34 how to perform basic arithmetic functions on the elements of one vector in R
```{r}
x<-1:5
x+10
x*3

```

0:04:02 how to preform arithmetic functions on the corresponding elements of two vectors in R
```{r}
y<-c(2,3,4,7,10)
x+y
x*y

```

0:04:59 how to extract elements of a vector using square brackets[] in R
```{r}
y
y[3]
y[-3]
y[1:3]
y[c(1,5)]
y[-c(1,5)]
y[y>5]
```

0:06:08 how to create a matrix using "matrix" command in R
```{r}
matrix(c(1,2,3,4,5,6,7,8,9),nrow=3,byrow=TRUE)
matrix(c(1,2,3,4,5,6,7,8,9),nrow=3,byrow=FALSE)
matrix(c(1,2,3,4,5,6,7,8,9),nrow=3)
matrix(c(1,2,3,4,5,6,7,8,9,10),nrow=2)

mat <- matrix(c(1,2,3,4,5,6,7,8,9),nrow=3,byrow=TRUE)
```

0:06:21 how to set the number of rows and columns in a matrix using "nrow" and "byrow" commands in R

0:07:05 how to use the square brackets to extract certain elements from a matrix in R
```{r}
mat[1,2]
mat[3,2]
mat[c(1,3),2]
mat[,2]
```

0:08:06 how to perform element-wise arithmetic functions in a matrix in R
```{r}
mat^2
```

### 1.5 Import csv
```{r}
LungCapData<-read.delim(file.choose(),header=T)
LungCapData
```

### 1.6 Working with data in R
```{r}
dim(LungCapData)
names(LungCapData)
mode(LungCapData)
class(LungCapData)
summary(LungCapData)

head(LungCapData)
tail(LungCapData,10)
LungCapData[c(10,20,30),]
LungCapData[21:24,]
LungCapData[-c(4:722),]

```

### 1.7 Working with Data in R
```{r}
mean(LungCapData$Age)

attach(LungCapData)
max(Age)
class(Age)
class(Smoke)
levels(Gender)
```

```{r}
x <- c(0,1,1,1,0,0,0,0)
class(x)
summary(x)

x<-as.factor(x)
class(x)
summary(x)

```

### 1.8 subsetting Date with square brackets
```{r}
Age[11:14]
mean(Age[Gender=="female"])
mean(Age[Gender=="male"])
mean(Age)

MaleOver15 <- LungCapData[Gender=="male" & Age>15,]
summary(MaleOver15)
```
### 1.9 Logic 

```{r}
Age[1:5]

temp1 <- Age>15
temp1[1:5]

temp2 <- as.numeric(Age>15)
temp2[1:5]
```

0:01:55 how to use multiple logical statements within an R command to have a logical vector answering multiple questions (step by step example)
```{r}
FemSmoke <- Gender=="female" & Smoke=="yes"
FemSmoke[1:5]
```


0:02:50 how to attach vectors or matrices in a column-wise fashion using the "cbind" command or in a row-wise fashion using the "rbind" command in R

```{r}
MoreData <- cbind(LungCapData, FemSmoke)
MoreData[1:5,]

rm(list=ls())

```

### 1.10 Working Directory
```{r}
getwd()

```

### 1.12 installing packages
```{r}
help("install.packages")
install.packages("epiR")
library(epiR)
help(package=epiR)
remove.packages(epiR)

```

### 2.1 Bar Charts and Pie Charts


```{r}
table(Gender)
count<-table(Gender)
pt<-table(Gender)/725
barplot(count)
barplot(pt, main="Gender PCt", xlab="Gender", ylab="%", names.arg = c("F","M"))

```

Modifying Pie Chart in R:

```{r}
pie(count)
```

### 2.2 Boxplot
A boxplot is for summarizing the distribution of a numeric variable
```{r}
boxplot(LungCap)
quantile(LungCap,probs=c(0,0.25,0.5,0.75,1))

boxplot(LungCap~Gender,main="Gender Lung Cap", ylim=c(0,16), las=1)
```

### 2.3 Stratifed boxplots
relationship between a categorical variable and a numeric variable, within strata or groups defined by a third categorical variable
```{r}
AgeGroups<-cut(Age, breaks=c(0,13,15,17,25), labels = c("<13","14/15","16/17","18+"))
Age[1:5]
AgeGroups[1:5]
levels(AgeGroups)
```

The Smoking effiect is confounded with the age efffect
on average, smokers are older than non-smokers, and older children have bigger lung capacities

```{r}
boxplot(LungCap~Smoke)

boxplot(LungCap[Age>=18]~Smoke[Age>=18])

boxplot(LungCap~Smoke*AgeGroups,col=c(4,2))

```
### 2.3 Histgrams
for summarizing the distribution of a numeric variable
```{r}
hist(LungCap)
hist(LungCap, freq = F)
hist(LungCap, prob = T, ylim=c(0,0.20), breaks=seq(0,16,2), las=1)
lines(density(LungCap),col=2,lwd=3)
```
### 2.4 Stem and leaf plots
```{r}
femalelungecap<-LungCap[Gender=="female"]
stem(femalelungecap)
stem(femalelungecap,scale=2)
```

### 2.5 Stacked barcharts, clustered barcharts, mosaic plot
examing the relationship between 2 categorical variables
```{r}
table1 <- table(Smoke,Gender)
table1
barplot(table1)
barplot(table1,beside = T, legend.text=T, col=c(3,2))

mosaicplot(table1)
```
### 2.6 Scatterplots
examing the relationship between two numerical variables

Pearson's correlation is used to examine the strength of the linear relationship between the 2 numeric variables
```{r}
cor(Age,Height)
plot(Age, Height)
abline(lm(Height~Age), col=4, lwd=3)

lines(smooth.spline(Age, Height), lty=2, lwd=5, col=2)

```
### 2.7 numerical summary

```{r}
table(Smoke)
table(Smoke)/length(Smoke)

table(Smoke,Gender)

mean(LungCap, trim=0.10)

median(LungCap)
var(LungCap)
sd(LungCap)
sqrt(var(LungCap))
sd(LungCap)^2
range(LungCap)
quantile(LungCap,probs = c(0.10,0.50,0.90))

cor(LungCap,Age)
cor(LungCap,Age,method = "spearman")
cov(LungCap,Age)
var(LungCap,Age)

summary(LungCap)
```

###3.1 Binomial Distribution


0:00:10 introducing the binomial random variable used in this video and its characteristics
choose(n,k)
```{r}
choose(5,0)
choose(5,1)
choose(5,2)
choose(5,3)
choose(5,4)
choose(5,5)

```

X is binomially distributed with n=20 trials and p=1/6 probability of success X ~ Bin (n=20,p=1/6)

0:00:43 how to use the "dbinom" command in R to calculate probability for the binomial random variable 
**dbinom** is used to find values for the probability density function of X, f(x)

P(X=3)
```{r}
dbinom(x=3,size=20,prob=1/6)
choose(20,3)*(1/6)^3*(1-1/6)^(20-3)

```
P(x=0) & p(x=1) & p(x=2) & p(x=3)
```{r}
dbinom(x=0:3,size=20,prob=1/6)
dbinom(x=0:1,size=20,prob=1/6)
```

P(x<=3)
```{r}
sum(dbinom(x=0:3,size=20,prob=1/6))
sum(dbinom(x=0:1,size=20,prob=1/6))
```



0:02:45 how to calculate cumulative probabilities in R for the binomial random variable using the "pbinom" command
```{r}
pbinom(q=3,size = 20, prob=1/6,lower.tail = T)
```
 "dbinom" find values for the probability _density_ function of X, f(x)
 "pbinom" gives values for the probability _distribution_ function of X, F(x)
 "rbinom" take a _random_ sample from a binomial dist.

 "qbinom" find _quantiles_ for a binomial dis.
 
###3.2 Poisson Distribution
the number of **discrete events or occurrences** over a specified **interval or continumm**(time, length, distance, etc.)

lambda= # occurences / specified interval
expected value, mean, lambda the same things.

checkout line, lambda=10 customer per 15 min interval.
outcome, discrete outcomes, x=0,1,2,3...infinite


x follows a poisson distribution with a known rate of lambda=7
X ~ poisson(lambda=7)
```{r}
dpois(x=4, lambda = 7)

7^4*exp(-7)/factorial(4)

dpois(x=0:4, lambda = 7)
sum(dpois(x=0:4, lambda = 7))
ppois(q=4, lambda = 7,lower.tail = T)
```

###3.3 Normally Distribution
X is normally distributed with a known mean of 75 and standard deviation of 5 
P(X<=70)
```{r}
pnorm(q=70, mean=75, sd=5, lower.tail = T)
```
P(Z>1)
```{r}
pnorm(q=1,mean=0,sd=1, lower.tail = F)
```
```{r}
qnorm(p=0.25, mean=75, sd=5,lower.tail = T)
qnorm(p=0.5, mean=75, sd=5,lower.tail = T)
qnorm(p=0.75, mean=75, sd=5,lower.tail = T)
qnorm(p=0.90, mean=75, sd=5,lower.tail = T)
```
```{r}
x<-seq(55,95,0.5)
dens<-dnorm(x,mean=75,sd=5)
plot(x,dens)
plot(x,dens,type="l")
abline(v=75)
rand<- rnorm(n=40, mean=75, sd=5)
hist(rand)
```

### 3.4 t Distribution and t Scores
these can be used to find p-values or critical values for constructing confidence intervals for statistics that follow a t-distribution

t follows a t-dis, with mean=0, sd=1 and 25 degrees of freedom.
P(t>2.3)
```{r}
pt(q=2.3, df=25, lower.tail=F)

pt(q=2.3, df=25, lower.tail=F)+pt(q=-2.3, df=25, lower.tail=T)
pt(q=2.3, df=25, lower.tail=F)*2

```
find t value for 95% confidence

```{r}
qt(p=0.025, df=25, lower.tail = T)
```

### 4.1 one-sample t Test
```{r}
boxplot(LungCap)
summary(LungCap)
```
Ho: mu>=8, Ha:mu<8
one-sided 95% confidence interval for mu
```{r}
t.test(LungCap, mu=8, alternative = "less", conf.level = 0.95)
```
two-sided
```{r}
t.test(LungCap, mu=8, alt = "two.sided", conf = 0.95)
TEST<-t.test(LungCap, mu=8,  conf.level = 0.99)
TEST
TEST$p.value
TEST$conf.int
```
### 4.2 two-sample t Test
these are ways of examing the relationship between a numeric outcome variable (Y) and a categorical explanatory variable (X, with 2 levels).

Ho : mean lung cap of sokers = of non somkers
two-sided test
assume non-equal variance

```{r}
t.test(LungCap~Smoke, mu=0, alt = "two.sided", conf = 0.95, var.eq=F, paired=F)
#same default value
t.test(LungCap~Smoke)
```

```{r}
var(LungCap[Smoke=="yes"])

var(LungCap[Smoke=="no"])
```
Levene's test
Ho: population variance are equal



### 5.1 Linear Regression

Model the relationship between Age and Lung Capacity
```{r}
cor(Age, LungCap)
mod<-lm(LungCap~Age)
summary(mod)
attributes(mod)
mod$coefficients


plot(Age,LungCap)
abline(mod)

```

